### 2024-04-30 manuscript.Rmd
Organize code before publication.

# Introduction

Genome-wide association studies (GWAS) have to date identified over 70 genetic risk loci for Alzheimer's Disease (AD).[@jansen2019; @wightman2021; @bellenguez2022] Most AD GWAS have focused on clinically diagnosed AD; however, clinically diagnosed AD is increasingly recognized as a heterogeneous disease that can reflect multiple independent or co-occurring pathological processes. Aged individuals in research autopsy cohorts often have more than one proteinopathy present in their brains at time of death.[@karanth2020]. Vascular factors, especially small-vessel disease, also contributes to cognitive decline.

does not always correspond with AD neuropathology and vice-versa.[@nelson2012]

We hypothesized that investigating the genetic architecture of multiple neuropathological endophenotypes (NPE) would:

1.  identify novel genetic risk loci for individual NPE
2.  identify common sources of genetic risk (pleiotropy) between NPE
3.  provide evidence that the effects of clinical AD genetic risk loci may be mediated by specific NPE.

>>>>

# Title page

**Title**

Genome-wide association study of multiple neuropathology endophenotypes identifies novel risk loci and provides insights into genetic risk of dementia

**Authors**

Lincoln M.P. Shade (1), Yuriko Katsumata (1, 2), Erin L. Abner (2, 3), Steven A. Claas (1), Timothy J. Hohman (4), Shubhabrata Mukherjee (5), Margaret A. Pericak-Vance (6), Richard P. Mayeux (7), Lindsay A. Farrer (8), Gerard D. Schellenberg (9), Jonathan L. Haines (10), Walter A. Kukull (11), Kwagsik Nho (12), Andrew J. Saykin (12), David A. Bennett (13), Julie A. Schneider (14), Mark T. W. Ebbert (2, 15, 16), Peter T. Nelson (16, 17), David W. Fardo (1, 2), the National Alzheimer's Coordinating Center, and the Alzheimer's Disease Genetics Consortium

**Add Qi and Khine as coauthors**

**Author Affiliations**

(1) Department of Biostatistics, College of Public Health, University of Kentucky, Lexington, KY, USA;

(2) Sanders-Brown Center on Aging and Alzheimer's Disease Research Center, University of Kentucky, Lexington, KY, USA;

(3) Department of Epidemiology and Environmental Health, College of Public Health, University of Kentucky, Lexington, KY, USA;

(4) Vanderbilt Memory & Alzheimer's Center, Department of Neurology, Vanderbilt University Medical Center, Nashville, TN, USA;

(5) Department of Medicine, University of Washington, Seattle, WA, USA;

(6) John P. Hussman Institute for Human Genomics, University of Miami Miller School of Medicine, Miami, FL, USA;

(7) Department of Neurology, Columbia University, New York, NY, USA;

(8) Department of Medicine, Boston University, Boston, MA, USA;

(9) Penn Neurodegeneration Genomics Center, University of Pennsylvania Perelman School of Medicine, Philadelphia, PA, USA;

(10) Cleveland Institute for Computational Biology, Case Western Reserve University, Cleveland, OH, USA;

(11) Department of Epidemiology, University of Washington, Seattle, WA, USA;

(12) Department of Radiology & Imaging Sciences, Indiana University School of Medicine, Indianapolis, IN, USA;

(13) Rush Alzheimer's Disease Center and Department of Neurological Sciences, Rush Medical College, Chicago, IL, USA;

(14) Rush Alzheimer's Disease Center and Departments of Neurology and Pathology, Rush University Medical Center, Chicago, IL, USA;

(15) Division of Biomedical Informatics, Department of Internal Medicine, University of Kentucky, Lexington, KY, USA;

(16) Department of Neuroscience, University of Kentucky College of Medicine, Lexington, KY, USA;

(17) Department of Pathology and Laboratory Medicine, University of Kentucky, Lexington, KY, USA

**Corresponding Author**

David W. Fardo

Healthy Kentucky Research Building

760 Press Ave, Room 372

Lexington, KY 40536-0679

[david.fardo\@uky.edu](mailto:david.fardo@uky.edu)

859-218-2070

# Key words

GWAS, Alzheimer's disease, neuropathology, atherosclerosis, neurofibrillary tangles, cerebral amyloid angiopathy, TDP-43

# Main

Late-onset amnestic dementia diagnosed as Alzheimer's Disease (AD/LOAD) is increasingly recognized as a heterogeneous clinical syndrome that can reflect multiple underlying independent or co-occurring pathological processes that may have independent or pleiotropic genetic risk factors.[@farfel2016; @katsumata2022; @karanth2020] Because of the complex nature of amnestic dementia and the practical challenge of constructing large cohorts with harmonized pathology data, LOAD genome-wide association studies (GWAS) to date have primarily utilized clinical diagnosis or proxy phenotypes based on family history of dementia to help unravel the genetic etiology of disease.[@jansen2019; @wightman2021; @bellenguez2022; @kunkle2019] While these LOAD GWAS have been immensely successful at identifying dementia-associated genetic loci, with over 80 such loci identified thus far,[@bellenguez2022] the clinical phenotypes used in most studies complicate the interpretation of their results and belies a more complex reality of mixed neuropathologies being the norm in aged individuals.[@karanth2020; @escott-price2022] To build on the successes of previous LOAD GWAS, and to begin decrypting the complexities of this heterogeneous clinical syndrome, we believe GWAS using neuropathology endophenotypes (NPE) is an essential next step--not only to broadly understand LOAD's genetic etiology, but to identify alleles that are drive specific LOAD-associated pathologies.[@karran2022]

Neuritic plaques and neurofibrillary tangles (NFT) are present in the brains of the majority of people diagnosed with clinical LOAD, but \~20% of diagnosed do not have substantial AD neuropathologic change (ADNC), and another \~50% have both ADNC and comorbid non-AD pathologies.[@nelson2009; @karanth2020] For example, TAR DNA-binding protein 43-kDa (TDP-43) pathology is found in 40-50% of autopsied individuals with dementia and contributes to an amnestic syndrome, limbic-predominant age-related TDP-43 encephalopathy (LATE), with a characteristic pattern of neuropathological change called LATE-NC.[@karanth2020; @nelson2019; @nelson2022] Hippocampal sclerosis (HS) of aging, a pathology characterized by neuronal death, gliosis, and atrophy of the hippocampus beyond normal ranges based on levels of AD pathology, commonly co-occurs with LATE-NC and is associated with severe cognitive impairment.[@brenowitz2014; @nelson2019] It is already known that the genetic risk factors of ADNC and LATE-NC differ,[@dugan2021] but the genetic underpinnings of common comorbid (ADNC+LATE-NC) phenotypes are not well understood.

Cerebrovascular pathologies, which contribute to cognitive decline and dementia, are also common in elderly autopsied research participants.[@skrobot2016] Cerebral amyloid angiopathy (CAA) is a cerebrovascular pathology characterized by amyloid-beta deposition in cerebral blood vessels.[@weber2018] CAA often co-occurs with LOAD but can independently contribute to cerebral injury by causing hemorrhages in the brain parenchyma.[@weber2018; @skrobot2016] Infarcts of both small but grossly visible lacunar arteries and microscopically examined vessels (the latter referred to as microinfarcts) are also common in aged individuals and contribute to cognitive decline.[@smith2012; @skrobot2016] Cerebral atherosclerosis and sclerosis of small cerebral blood vessels, called brain arteriolosclerosis, predispose individuals to infarcts and HS [@arvanitakis2017; @neltner2014]; moreover, they contribute to cognitive decline even after adjusting for the presence of related pathologies.[@arvanitakis2016; @ighodaro2017] Collectively, these factors reveal an increasingly complex and synergistic web of pathologies which contribute to cognitive impairment and dementia and further demonstrate the need to understand how existing (and potentially new) GWAS loci are involved in LOAD-associated pathologies.

Studying the genomic risk factors of individual underlying neuropathologies can provide a complementary approach to large GWAS of clinical- and family history-based outcomes for the study of AD and dementia risk. As proof of principle, prior GWAS of neuropathology endophenotypes (NPE) have confirmed known LOAD risk loci and have identified potential new neuropathology risk loci.[@beecham2014; @shade2022; @vattathil2021; @farrell2022; @reddy2021] Some NPE, particularly LATE-NC, have yet to be studied using GWAS; statistical methodological development for GWAS and downstream functional analyses along with continued recruitment in autopsy cohorts provide additional opportunity for in-depth investigation into multiple neuropathologies. Here, we perform GWAS on eleven neuropathology endophenotypes in the largest harmonized cohort, to date, using three high-quality data sources with autopsy and genotype data. For previously studied NPE, the sample sizes used here represent 50-150% increases relative to those used in prior studies. We also perform downstream functional analyses to explore potential biological functional mechanisms of newly identified risk loci and provide insight into known AD risk loci.

## Participants and distributions of neuropathology endophenotypes

```{r}
add_comma <- function(n) {
  formatC(n, big.mark = ",")
}
nacc_covar <- fread("data/adc/adc_np.covar")
rosmap_covar = fread("data/rosmap/rosmap_np.covar")
act_covar = fread("data/act/act_np.covar")
covar_vars = c("FID", "IID", "msex", "age_death", paste0("pc", 1:10))
covar = rbindlist(
  list(
    nacc_covar[, ..covar_vars], 
    rosmap_covar[, ..covar_vars], 
    act_covar[, ..covar_vars]), 
  use.names = TRUE
)
np_pheno <- fread("shared/nacc_rosmap_act_np.pheno")
nmega <- formatC(np_pheno[, .N], big.mark = ",")
nnacc <- formatC(np_pheno[FID == "0", .N], big.mark = ",")
nrosmap <- formatC(np_pheno[FID == "1", .N], big.mark = ",")
nact <- formatC(np_pheno[FID == "2", .N], big.mark = ",")
# nadni <- formatC(covar[FID == "ADNI", .N], big.mark = ",")

# np_pheno_ord <- fread("data/mega/mega_np_ord.pheno", na.strings = "-1")
# np_pheno_bin <- fread("data/mega/mega_np.pheno", na.strings = "-1")

# rm(np_pheno_ord, np_pheno_bin)
ncerad <- np_pheno[!is.na(cerad), add_comma(.N)]
nbraak <- np_pheno[!is.na(braak), add_comma(.N)]
# ndiffuse_abeta <- np_pheno[!is.na(diffuse_abeta), add_comma(.N)]
narteriol <- np_pheno[!is.na(arteriol), add_comma(.N)]
natheroscler <- np_pheno[!is.na(atheroscler), add_comma(.N)]
ncaa <- np_pheno[!is.na(caa), add_comma(.N)]
nlate <- np_pheno[!is.na(late), add_comma(.N)]
nlewy = np_pheno[!is.na(lewy_body), add_comma(.N)]
nhs = np_pheno[!is.na(hs), add_comma(.N)]
nmicroinf = np_pheno[!is.na(microinf), add_comma(.N)]
ngrossinf = np_pheno[!is.na(grossinf), add_comma(.N)]
```

We used genotype and neuropathology data from three autopsy data sources: (1) the National Alzheimer's Coordinating Center (NACC, N = `r nnacc`), (2) the Religious Orders Study and the Rush Memory and Aging Project (ROSMAP, N = `r nrosmap`), and (3) the Adult Changes in Thought (ACT, N = `r nact`) study. In total, `r nmega` unique participants were included in our analyses. NACC participants were more likely to be diagnosed with either dementia (83%) or AD/mild cognitive impairment (MCI, 77.5%) than ROSMAP (55% and 66%) or ACT (13% and 10%) participants. NACC participants died at a younger age (mean age at death 81 years) compared to ROSMAP (90 years) and ACT (89 years) participants and had more balanced participation between the sexes, with 50% of NACC participants being female vs. 68% and 58% in ROSMAP and ACT, respectively. NACC participants were also more likely to have an *APOE* $\epsilon 4$ allele (54%) vs. ROSMAP (25%) or ACT (28%; Chi-square test $P \text{-value}<0.001$ for all comparisons). These differences reflect heterogeneity in the recruitment for these studies. NACC is based on data collected from over 30 Alzheimer's Disease Research Centers which often recruit from clinic patients and their families. In contrast, ACT originally recruited aged participants in the Seattle, WA area who did not have dementia at time of enrollment, and ROS recruited members of the Catholic Church clergy.

Available sample sizes differed among the eleven studied NPE, which included the AD-related pathologies of Consortium to Establish a Registry for Alzheimer's Disease (CERAD) score for neuritic plaques (\$N = `r ncerad`, Braak staging of NFT (N = `r nbraak`), and diffuse amyloid-beta plaques; the non-AD proteinopathies LATE-NC (N = `r nlate`) and Lewy body pathology (N = `nlewy`); hippocampal sclerosis (N = `r nhs`); and the cerebrovascular pathologies CAA (N = `r ncaa`), gross infarcts (N = `r ngrossinf`), microinfarcts (N = `r nmicroinf`), atherosclerosis (N = `r natheroscler`), and arteriolosclerosis (N = `r narteriol`). Relative to the Beecham et al. (2014) GWAS of neuropathology endophenotypes, our updated sample represents an overall 52% increase in sample size.[@beecham2014] The sample size increase for individual neuropathologies is greater, ranging from 57% for Braak stage to 150% for CAA, demonstrating both an increased total number of autopsy participants and more complete neuropathological profiling for each participant. HS, microinfarcts, and gross infarcts were coded as binary case/control outcomes. All other phenotypes were ordinal variables with either four or seven levels (Methods and Supplementary Note describe our phenotype definitions and harmonization approach).

To explore the co-occurrence of neuropathology endophenotypes in our pooled data, we estimated polychoric correlations (an approach which assumes that observed ordinal or binary variables reflect latent normally distributed variables) between neuropathology endophenotype pairs followed by hierarchical clustering using the polycor, psych, and pheatmap R packages.[@revelle2022; @fox2022; @kolde2019] Inspection of the resulting dendrogram identified three positively correlated clusters of endophenotypes: a "vascular" cluster consisting of gross infarcts, microinfarcts, arteriolosclerosis, and atherosclerosis; an "Alzheimer's disease" cluster consisting of Braak NFT stage, neuritic plaques, diffuse amyloid-beta plaques, and CAA; and a "LATE" cluster consisting of LATE-NC and HS (Supplementary Figure 1).

## GWAS meta-analysis of eleven neuropathology endophenotypes (NPE)

An overview of our study design is shown in **Figure 1**. Genetic association analyses were performed with logistic or proportional-odds logistic regression mixed-effects models as appropriate (Methods). We performed GWAS on 11 NPE using NACC participants, ROSMAP, and ACT separately. We then performed fixed-effect meta-analysis using METAL on variants with minor allele frequencies $\ge 1\%$ in each data source.[@metal:f]

```{r}
t2_env <- new.env()
load("output/manuscript/table2_meta_analysis_results.Rdata",
       envir = t2_env)
```

In total, 13 NPE-genomic locus associations had lead variants that met the genome-wide significance threshold of $P \text{-value}< 5 \times 10^{-8}$. Three NPE (Braak stage, CERAD score, and CAA) were associated with two or more independent loci ($r^2\le0.05$) in the *APOE* region, though we report only one locus in the *APOE* region for each phenotype unless secondary loci remained significant after conditioning on *APOE* $\epsilon$ diplotype (see conditional *APOE*-adjusted analysis below). The lead variant rs429358 in the *APOE* region was significantly associated with four NPE: Braak stage ($\text{OR} = 2.06$, $P \text{-value}= 9.7\times 10^{-89}$), CAA ($\text{OR} = 2.49$, $P \text{-value}= 4.4\times 10^{-138}$), CERAD score ($\text{OR} = 2.42$, $P \text{-value}= 4.7\times 10^{-103}$), and LATE-NC ( $\text{OR} = 1.70$, $P \text{-value}= 1.7\times 10^{-14}$). Variants within *TMEM106B* were associated with both HS (rs7805419; $\text{OR} = 0.65$, $P \text{-value}= 3.2\times 10^{-13}$) and LATE-NC (2043539; $\text{OR} = 0.70$, $P \text{-value}= 5.8\times 10^{-11}$). A *GRN* locus was associated with HS (rs5848; $\text{OR} = 1.40$, $P \text{-value}= 3.2\times 10^{-8}$). One *BIN1* locus was associated with Braak stage (rs6733839 $\text{OR} = 1.21$, $P \text{-value}= 1.6\times 10^{-9}$).

Novel associations identified in the NPE GWAS meta-analysis included a variant 12 kbp upstream of *COL4A1* associated with cerebral atherosclerosis (rs2000660; $\text{OR} = 0.73$, $P \text{-value}= 2.7\times 10^{-8}$), a *PIK3R5* intronic locus associated with Braak stage (rs72844606 $\text{OR} = 0.69$, $P \text{-value}= 4.0\times 10^{-8}$), and an intronic *LZTS1* locus associated with arteriolosclerosis (rs78909048; $\text{OR} = 0.44$, $P \text{-value}= 5.7\times 10^{-10}$). One intronic locus of *SPATA48* was close to genome-wide signicantly associated with atherosclerosis (rs62447817; $\text{OR} = 1.35$, $P \text{-value}= 5.5\times 10^{-8}$)

## Conditional *APOE*-adjusted analysis

To check whether any loci in the *APOE* region (defined as less than 500 kbp from the start or end site of *APOE* transcription) were associated with NPE risk independently of the known effects of *APOE* $\epsilon 2/\epsilon 3/\epsilon 4$ alleles, we re-analyzed the *APOE* region of Chromosome 19 while adjusting for *APOE* $\epsilon$ diplotypes. We limited this conditional analysis to NPE with which variants within the *APOE* region met the genome-wide significance threshold of $P \text{-value}< 5\times 10^{-8}$ in Stage 3 (Braak NFT stage, neuritic plaque CERAD score, diffuse amyloid-beta plaques, cerebral amyloid angiopathy, and LATE-NC). One locus between *APOC4-APOC2* with lead variant rs7247551 remained significantly associated with CAA after adjusting for *APOE* $\epsilon 2/\epsilon 3/\epsilon 4$ diplotypes ($\text{OR} = 0.81$, $P \text{-value}= 8.0\times 10^{-12}$, Table 1 and Figure 3a). No variants remained genome-wide significantly associated with any other APOE-associated NPE, and the P-values of the lead variant rs429358 from genome-wide meta-analysis unadjusted for APOE were all above 0.05 in the re-analysis. Using independent data from a recent GWAS of CAA in 815 participants with dementia in the Mayo Clinic Brain Bank, we replicated the association of rs7247551 with CAA while adjusting for APOE $\epsilon$ diplotype ($\beta = -0.068$, $P \text{-value} = 0.0012$). These results provide evidence for a novel locus in the *APOE* region other than the known *APOE* $\epsilon$ haplotypes that affects CAA pathology burden, and that the genetic risk basis of CAA in the *APOE* region may differ from that AD neuropathologies of neuritic plaques and NFT.

## Associations of clinical and proxy AD risk loci with NPE

We checked whether AD-associated loci identified in the recent Bellenguez et al. (2022) GWAS of AD and related dementia (ADD) were associated with NPE.[@bellenguez2022] Bellenguez et al. identified a total of 83 non-*APOE* independent loci (39 previously identified, 44 novel) associated with ADD (hereafter, "ADD loci"), 80 of which had lead variants that were included in our study. Each of the 11 NPE used as outcomes in the present study was checked for association with variants in these 80 ADD loci. Benjamini-Hochberg adjustments were then applied to *P*-values for each NPE separately to control the false-discovery rate (FDR). In total, 22 NPE-locus associations had adjusted *P*-values (*Q*-values) $\le 0.05$ and were considered significant; of these 22 associations, 21 had concordant directions of effect with Bellenguez et al. (Table 2 and Supplementary Table 5). The pathognomonic AD pathologies, Braak stage and CERAD, had 70 and 60 of 80 concordant directions of effect with ADD loci, respectively, regardless of statistical significance. Three ADD loci (*BIN1, MME,* and *HLA*) were significantly associated with Braak NFT stage after multiple test correction. Ten AD loci (*CR1*, *BIN1, INPP5D,* ZCWPW1/NYAP1*, PTK2B, PICALM*, *SORL1*, *FERMT2*, *SNX1,* andABCA7) were significantly associated with CERAD score after multiple test correction. Four AD loci (*TMEM106B, IL34, GRN*, and *MAPT*) were significantly associated with HS, all of which except for *L34* (HS $\text{OR} = 0.74$, $P \text{-value}=0.0019$; AD $\text{OR} = 1.06$, $P \text{-value}=5.6\times 10^{-6}$) were concordant in effect direction. Two ADD loci (*TMEM106B* and *GRN*) were significantly associated with LATE-NC, both concordant. Finally, one ADD locus (*PLCG2*) was significantly associated with microinfarcts. These results indicate that NPE studies largely corroborate the findings of large AD GWAS based on clinical and proxy phenotypes. However, several ADD loci, particularly *TMEM106B* and *GRN*, are associated with non-AD pathology (HS and LATE-NC) but not neuritic plaques or NFT.

```{r}
load("output/manuscript/results_ad_snps_table.Rdata")
results_ad_snps_table
```

## Gene-prioritization and enrichment analyses

We used gene-based, pathway, and enrichment analyses implemented in FUMA to identify potential functional genes and biological pathways involved in neuropathology development (results available in Supplementary Data zip file).[@watanabe2017] We used a minimum threshold of association of $P \text{-value}\le 1 \times 10^{-5}$ in the Stage 3 analysis for variant annotation. Gene-level analyses implemented in MAGMA corroborated our single-variant analyses,[@leeuw2015] identifying gene associations with neuropathology endophenotypes in regions with significant variant-phenotype associations. *APOE* was significantly associated with NFT, diffuse plaques, CAA, neuritic plaques, and LATE-NC. *TMEM106B* was associated with both HS and LATE-NC. Gene-set enrichment analyses were performed using MAGMA and Gene Ontology (GO) and curated gene sets from MSigDB.[@liberzon2011] After Bonferroni correction genes involved in polysaccharide digestion were significantly associated with microinfarcts ($P \text{-value}_{\text{bon}} = 5.3\times10^{-5}$). We also found evidence that genes involved in substrate recognition for ubiquination were associated with cerebral atherosclerosis ($P \text{-value}_{\text{bon}} = 0.0054$), and genes overexpressed in developing B cells were associated with Braak NFT stage ($P \text{-value}_{\text{bon}} = 0.024$). FUMA variant annotation using CADD score, RegulomeDB score, eQTL, and chromatin interaction annotations indicated that the rs2000660 lead variant associated with cerebral atherosclerosis in Stage 3 is located within an enhancer region located 13 Kbp upstream of *COL4A1* transcription start site.

## Colocalization analysis

We investigated whether loci associated with multiple NPE colocalized using a Bayesian colocalization analysis approach implemented in the coloc R package.[@giambartolomei2014] Because CAA was the only NPE with an associated locus in the *APOE* region with a signal independent of *APOE* $\epsilon$ diplotype, we excluded this region from NPE-NPE colocalization analysis. Pairs of NPE with the same variant reaching a P-value threshold of $P \text{-value}< 1 \times 10^{-5}$ with concordant effect directions included (1) Braak NFT stage and CERAD score in the *BIN1* region on Chromosome 2 and (2) HS and LATE-NC in the *TMEM106B* region on Chromosome 7 (**Figure 4**). Both loci exhibited high posterior probability of colocalization ($\text{PrC}> 90\%$), indicating that the same genetic variants drive association with both endophenotypes.

A locus within *GRN* associated with HS ($P \text{-value}= 3.2 \times 10^{-8}$) strongly colocalized with *GRN* expression in multiple tissues ($\text{PrC}>99.9\%$). The *TMEM106B* locus associated with HS and LATE-NC colocalized with *TMEM106B* expression in multiple tissues, including the cerebellar hemisphere ($\text{PrC}=90\%$). Two CpG sites located either within *TMEM106B* (cg09613507) or downstream (cg23422036) colocalized with both HS and late (cg09613507-HS $\text{PrC}=94\%$, cg09613507-LATE-NC $\text{PrC}=89\%$, cg23422036-HS $\text{PrC}=94\%$, cg23422036-LATE-NC $\text{PrC}=95\%$).

The Chromosome 19 CAA-associated locus identified in the *APOE*-adjusted analysis colocalized with expression of multiple genes in GTEx: *APOE* expression in the wall of the aorta; *8CLPTM1* expression in the skin of the leg and suprapubic region; and *APOC2* expression in 14 different tissues, including the brain cortex, caudate, nucleus accumbens, and cerebellum ($\text{PrC}= 90\text{-}97\%$). The same locus colocalized with mQTL for four CpG sites in ROSMAP, cg09555818, cg04401876, cg10169327, and cg13119609 ($\text{PrC}= 96\text{-}99\%$, see **Figure 3b**).

Multiple suggestive NPE loci showed evidence of colocalization with eQTL in GTEx. In total, 44 NPE loci (lead variant $P \text{-value} < 1 \times 10^{-5}$) colocalized with QTL with $\text{PrC}= \ge80\%$, with 533 total colocalizing NPE-QTL pairs across 44 tissues (Supplementary Table S4).

## Multiple eQTL and mQTL colocalizing with NPE are validated with methylation and RNA-seq in ROSMAP

Using ROSMAP participants with available DNA methylation and/or RNA-Seq data available from the DLPFC, we further investigated associations between QTL phenotypes found to colocalize with NPE in brain tissues of GTEx and/or ROSMAP. We tested methylation levels at each of the four CpG sites -- cg09555818, cg04401876, cg10169327, and cg13119609 -- that colocalized with the Chromosome 19 CAA risk locus for association with to confirm association with lead variant rs7247551. We first confirmed that all four CpG sites were significantly associated ($P \text{-value} < 0.0001$) with rs7247551 with directions of effect consistent with those previously reported for ROSMAP.[@ng2017] We then tested for association between methylation at these sites with CAA pathology. Hypomethylation at cg09555818 (odds ratio $[\text{OR}] =0.82$, $P \text{-value}= 0.004$) and cg13119609 ($\text{OR} =0.78$, $P \text{-value}= 0.0007$) were significantly associated with worse CAA pathology (Figure 3C).

Next, as *APOC2* expression in multiple brain tissues colocalized with CAA in GTEx but not ROSMAP, we investigated to see whether there was a nominal association between *APOC2* expression in the DLPFC and rs7247551. We found that the G allele of rs7247551 was significantly associated with increased *APOC2* expression in the DLPFC ($\beta = 0.072$, $P \text{-value}= 0.00015$, Figure 3D); however, the direction of effect was opposite of that found in brain tissues in GTEx (i.e., the G allele of rs7247551 was associated with decreased *APOC2* expression in GTEx, Figure 3E). Expression of *APOC2* in the DLPFC was not associated with CAA in ROSMAP ($\text{OR} =0.98$, $P \text{-value}= 0.89$). We performed an additional *post-hoc* analysis for nominal *APOE* eQTL activity of rs7247551 in ROSMAP (rs7247551 was lead eQTL in colocalizing *APOE* expression loci in the aorta, thyroid, and prostate, though APOE expression in no brain tissues colocalized with CAA). We confirmed that rs7247551 was not associated with *APOE* expression in the DLPFC in ROSMAP ($P \text{-value}= 0.75$, Figure 3D).

We then investigated the associations betwen HS and LATE-NC with the *GRN* and *TMEM106B* expression and methylation in the DLPFC. We first confirmed that the lead *TMEM106B* variants for HS (rs7805419) and LATE-NC (rs2043539) were in LD ($R^2 = 0.96$) and that rs2043539 was associated with DNA methylation levels at cg09613507 and cg23422036 as previously reported. We also found that Neither TMEM106B nor GR*N* expression were associated with HS ($P \text{-values} > 0.05$), while decreased *TMEM106B* expression was associated with more severe LATE-NC pathology ($P \text{-value}= 0.035$). Of the two CpG sites that colocalized with HS and LATE-NC, hypermethylation of cg09613507 was associated with more severe LATE-NC pathology ($P \text{-value}= 0.0093$).

The present study of eleven neuropathology endophenotypes with a maximum sample size of 7,786 participants serves as an autopsy-based complement to previous studies, expanding on the findings of previous genetic association studies of NPE, Alzheimer's, and related dementias to attempt to better understand the complex associations between different neuropathologies and genetic risk factors.[@katsumata2022; @shade2022; @beecham2014; @dugan2021; @bellenguez2022; @vattathil2021; @reddy2021] We identified a novel *APOE* $\epsilon$-independent CAA risk locus that also affects *APOC2* brain expression and dementia-associated CpG methylation sites. Two of these CpG sites were in turn also associated with CAA risk. We also discovered intriguing new loci mapped to *COL4A1*, *PIK3R5*, and *LZTS1* associated with atherosclerosis in the circle of Willis, Braak stage for neurofibrillary tangles, and brain arteriolosclerosis, respectively. Lastly, we investigated known loci in *BIN1*, *APOE*, and *TMEM106B* to provide additional context on their associations with multiple NPE.

Multiple variants in the *APOE* region were associated with pathognomonic AD NPE, including Braak stage, CERAD score, diffuse amyloid plaques, and CAA. Variants in the *APOE* region were also associated with LATE-NC, which is consistent with previous genetic association studies of NPE.[@beecham2014; @dugan2021] Notably, *APOE* variants were not significantly associated with vascular pathology except for CAA. When we re-ran Chromosome 19 analyses on these phenotypes while adjusting for *APOE* $\epsilon$ genotype, only one CAA-associated locus with lead variant rs7247551 remained significant ($\text{OR} = 0.81$, $P \text{-value}= 8.0 \times 10^{-12}$). We confirmed that this locus was indeed novel and not in LD with a recent study of CAA in *APOE* $\epsilon 4$ carriers by Reddy et al. (2021) by confirming that $R^2 <0.01$ between rs7247551 and Reddy et al. lead variant rs5117.[@reddy2021] Futhermore, we were able to replicate the associated between rs7247551 and CAA pathology while adjusting for *APOE* $\epsilon$ diplotype in 815 Mayo Clinic Brain Bank participants used in Reddy et al. but not used in the present CAA GWAS ($\beta = -0.068$, $P \text{-value} = 0.0012$), providing additional evidence for the APOE locus being important for CAA pathology beyond the known effects of APOE $\epsilon$ haplotypes.

Several variants in the novel CAA locus were lead eQTL for *APOC2* and brain expression in GTEx Colocalization analysis confirmed that the CAA risk locus shares a functional variant with both *APOC2* eQTL ($\text{PrC} = 89-97\%$ in brain tissues) and nearby mQTL ($\text{PrC} = 97\text{-}99\%$). This locus also colocalized with *APOE* expression in the wall of the aorta ($\text{PrC} = 94\%$) and prostate ($\text{PrC} = 86\%$) but not in any brain tissues. Furthermore, a lead QTL in this locus, rs4803779, was associated with *APOC2* expression but not *APOE* expression in the DLPFC in ROSMAP.

Variants in the same locus were also mQTL in the DLPFC in ROSMAP for four methylation sites (rs4803779 $P \text{-value}= 3.3 \times 10^{-91}$ for top site cg09555818; Figure 3). We confirmed that two of the CpG sites affected by the CAA risk locus, cg09555818 ($\text{OR} =0.82$, $P \text{-value}= 0.004$) and cg13119609 ($\text{OR} =0.78$, $P \text{-value}= 0.0007$), were in turn significantly associated with CAA pathology. Both of these CpG sites are located within the *APOC4-APOC2* readthrough transcript region overlapping *APOC4* and *APOC2*. Notably, we found that methylation levels at cg09555818 and cg13119609 are highly positively correlated ($\rho = 0.90$, $P \text{-value} < 2.2 \times 10^{-16}$) in ROSMAP. Our results are consistent with the hypothesis that rs7247551 locus affects CAA risk through hypomethylation of CpG sites in the *APOC4-APOC2* region. These results are also consistent with previous studies in other human cohorts that implicate hypomethylation at cg09555818 and cg13119609 as potentially associated with Alzheimer's disease.[@walker2021; @walker2020; @shao2018] Collectively, these results provide circumstantial evidence indicates that *APOC2* may be the target gene of the rs7247551 CAA risk locus as it colocalizes with *APOC2* expression in multiple brain tissues in GTEx and the associated CpG sites are located in exon 3 of *APOC4-APOC2*. However, more research must be done to verify that *APOC2* is the target gene.

One intronic locus of *TMEM106B* was significantly associated with both HS ($\text{OR} = 1.50$, $P \text{-value}= 1.5 \times 10^{-11}$) and LATE-NC ($\text{OR}= 1.50$, $P \text{-value}= 1.5 \times 10^{-11}$). Additionally, a locus within *GRN* was associated with HS ($\text{OR} = 0.72$, $P \text{-value}= 3.8 \times 10^{-8}$). Both of these genes are associated with frontotemporal lobar degeneration,[@rollinson2011; @ciani2019] HS,[@katsumata2017; @nelson2015] and ADD.[@bellenguez2022; @wightman2021] We found that HS, LATE-NC, and ADD all colocalize at these two loci ($\text{PrC}\ge 90\%$, Figure 5A). These results indicate that HS, LATE-NC, and LOAD likely share causal loci for these genes. When NPE were regressed on known AD variants, Braak NFT stage was significantly associated with *TMEM106B* with a direction of effect opposite of both recent AD GWAS and other neuropathologies,[@bellenguez2022; @wightman2021] while neuritic plaques were not associated with *TMEM106B* in our study. Evidence from RNA-Seq analyses supported the hypothesis that *TMEM106B* may have discordant effects on TDP-43 and NFT pathology, and both the genetic and gene expression associations between *TMEM106B* and NFT were strengthened when adjusting for LATE-NC stage. While preliminary, these results support the possibility that the associations between *TMEM106B* and AD found in recent GWAS are driven by non-AD pathologies, namely TDP-43, in dementia cases, and may mask a more complex association between *TMEM106B* and NFT.

A locus approximately 30 kbp downstream of *BIN1* on Chromosome 2q14 was significantly associated with Braak stage ($\text{OR} = 1.21$, $P \text{-value}= 3.8 \times 10^{-10}$) and suggestively associated with CERAD score for neuritic plaques ($\text{OR} = 1.20$, $P \text{-value}= 6.6 \times 10^{-8}$). We verified through colocalization analysis that the same locus drives association signals with Braak NFT stage and CERAD score. In prior GWAS, this locus is second only to *APOE* for strength of association with LOAD.[@bellenguez2022] Interestingly, the lead variant in this locus, rs6733839, was associated with neither diffuse amyloid plaques ($\text{OR} = 1.05$, $P \text{-value}= 0.15$) nor CAA ($\text{OR} = 1.02$, $P \text{-value}= 0.49$). Previous research supports the hypothesis that *BIN1* is associated with LOAD through its effect on NFT rather than amyloid pathology.[@holler2014; @franzmeier2019] As neuritic plaques contain dying nerve cell processes with aberrant tau fibrils identical to those seen in neurofibrillary tangles,[@nelson2009] our findings are also consistent with the hypothesis that *BIN1* influences AD risk primarily through tau rather than amyloid pathogenic processes.
